<<<<<<< HEAD
---
title: "College Voting"
output:
  pdf_document: default
  html_notebook: default
---

```{r}
#Load the necessary libraries
library(ipumsr)
library(tidyverse)
library(readxl)
library(scales)
```


```{r}
#Read in data file and format to grid view
ddi <- read_ipums_ddi("C:/Users/Rinija/Documents/Other/Projects/college-voting/TiS FDC 2020/TiS FDC 2020/data_fixed_width_for_R/cps_00008.xml")
data <- read_ipums_micro(ddi)
View(data)
```

```{r, eval=FALSE}
#Filter data so that only people that are registered to vote included
#Filter data so that only people who "Voted" or "Did not vote" were included
data <- data %>%
  filter(is.na(VOYNOTREG) | VOYNOTREG != 8)  %>%
  filter(VOTED == 1 | VOTED == 2 ) %>%
  filter(!(VOTED == 1 & VOREG == 98))
```

```{r}
#Remove NA entries and re-factor the categorical values  
data <- data %>%
  
  #Metro data into 3 categories
  mutate(
    METROSIMPLE = case_when(
      METRO == 1 ~ 0,  #Not in metro area
      METRO == 2 ~ 1,  #Central city
      METRO == 3 ~ 2   #Outside central city
        )
    )%>% 
  #Get rid of NA values
  filter(!is.na(METROSIMPLE)) %>% 
  #Rename the metro categories
  mutate(METROSIMPLE = fct_recode(as.character(METROSIMPLE), 
                                "Non-metro area" = "0",
                                 "Central city" = "1", 
                                 "Outside central city" = "2"
                         
                            )) %>% 
  #Race data into 4 main categories and others are in the 5th category
  mutate(
    RACESIMPLE = case_when(
      RACE == 100 ~ 1, #White
      RACE == 200 ~ 2, #Black
      RACE == 300 ~ 3, #American Indian/Aleut/Eskimo
      RACE == 651 ~ 4, #Asian only
      RACE > 651 ~ 5,  #Other
      RACE < 830 ~ 5   #Unspecified
         )
     ) %>% 
  #Rename the race categories
  mutate(RACESIMPLE = recode_factor(as.character(RACESIMPLE), 
                                "1" = "White",
                                 "2" = "Black", 
                                 "3" = "American Indian",
                                "4" = "Asian",
                                "5" = "others" )) %>% 
  #Remove NIU variable
  filter(SEX == 1 | #male
         SEX == 2   #female
         ) %>% 
  #Rename sex categories
  mutate(SEXSIMPLE = recode_factor(as.character(SEX), "1" = "Male","2" = "Female")) %>% 
  
  #Re-factor 
  mutate(
    VOTEDSIMPLE = case_when(
      VOTED == 1 ~ 0, #Did not vote
      VOTED == 2 ~ 1  #Voted
        )
  )  %>% 
  
  #Only isolate college students
  filter(AGE > 17, AGE < 23) %>% 
  filter(!is.na(VOTEDSIMPLE)) %>% 
  mutate(VOTEDSIMPLE = recode_factor(VOTEDSIMPLE, "0" = "No Vote","1" = "Vote")
         )
```

Preliminary Analysis of Variable Distributions:

Independent & Dependent Variables ->
  Y-variable - Voted
  X-variables - Age, Metro, Race, SEX
```{r}
#Preliminary Analysis of VOTEHOWSIMPLE
summary(data$VOTEDSIMPLE)
#hist(data$VOTEDSIMPLE)
table(data$VOTEDSIMPLE)
```
VOTEDSIMPLE Data Analysis:

The majority of the population voted and the second highest concentration did not vote. There is roughly half as many people that did not vote when compared to the people that did vote. 
  
```{r}
#Preliminary Analysis of AGE
summary(data$AGE)
hist(data$AGE)
boxplot(data$AGE)
```
Age Data Analysis:

The highest concentration of people is between ages of 55-60
It is worth noting that there were fewer people in this study that were between the ages of 18-20 and ages 70-85. 

```{r}
#Preliminary Analysis of METRO
summary(data$METROSIMPLE)
table(data$METROSIMPLE)
```
Metro Data Analysis:

The majority of people in this data set live outside central city and the second highest category is people that live in central cities.


```{r}
#Preliminary Analysis of Race
summary(data$RACE)
hist(data$RACE)
table(data$RACE)
```
Race Data Analysis:

The majority of people in this data set are white. The number of minorities is much smaller compared the the number of white voters. 

```{r}
#Preliminary Analysis of Veterans
summary(data$VETSTAT)
hist(data$VETSTAT)
table(data$VETSTAT)
```
Veterans Data Analysis:

The majority of people in this data set did not serve veterans and only 0.102434 people served.  

```{r}
#Plot voting rates in terms of where people lived 
data %>% select(VOTEDSIMPLE,AGE,RACE,METROSIMPLE,SEXSIMPLE) %>% 
 
  ggplot()+
  geom_bar(aes(x=VOTEDSIMPLE))+
  
  facet_wrap(~METROSIMPLE)+
  scale_y_continuous(label = scientific_format(digits = 2))+
  labs(title = "Voting Rate Based on Location")
```



```{r}
#Plot voting rates in terms of people's race
data %>% 
  select(VOTEDSIMPLE,AGE,RACESIMPLE,METROSIMPLE,SEXSIMPLE) %>% 
  ggplot()+
  
  geom_bar(aes(x=VOTEDSIMPLE)) +
  facet_grid(~RACESIMPLE)+

  labs(title = "Voting Rate Based on Race")
```



```{r}
#Plot voting rates in terms of sex
data %>% 
  select(VOTEDSIMPLE,AGE,RACE,METROSIMPLE,SEXSIMPLE) %>% 
 
  ggplot()+
  geom_bar(aes(x=SEXSIMPLE, fill=VOTEDSIMPLE))+
  labs(title = "Voting Rate Based on Sex")
```



```{r}
#Plot voting rate depending on age of college students
data %>% 
  select(VOTEDSIMPLE,AGE,RACE,METROSIMPLE,SEXSIMPLE) %>% 

  ggplot()+
  geom_bar(aes(x=AGE, fill=VOTEDSIMPLE))+
  labs(title = "Voting Rate Based on Age")
```



Logistic Regression:

```{r}
summary(glm(VOTEDSIMPLE ~ AGE + METROSIMPLE + RACESIMPLE + SEXSIMPLE, family = binomial(link = "logit"), data = data))
```
Interpretation of Logistic Regression Model:

Assumptions: Each variable does not effect the other ones. This means that race will not have an effect on metro, sex, or age. 

Model equation:
  log_odds =  -2.164897 + (0.077789*AGE) + (0.213431*METROSIMPLECentralCity) + (0.215372*METROSIMPLEOutsideCentralCity) + (0.165660*RACESIMPLEBlack) + (-0.542498*RACESIMPLEAmericanIndian) + (-0.354319*RACESIMPLEAsian) + (-0.168815*RACESIMPLEOther)+ (0.219261*SEXSIMPLEFemale)

Hypothesis Test:

  Null Hypothesis -> AGE, RACE, METRO, and SEX did not impact VOTED
  Alternate Hypothesis -> Either AGE, RACE, METRO, or SEX impacts VOTED
  
  Since each variable has a p-value that is less than 0.05, each variable has a significant impact on whether that person votes or not.
  
Values in transformed data:
```{r}
summary(data$AGE)
summary(data$METROSIMPLE)
summary(data$VOTEDSIMPLE)
summary(data$RACESIMPLE)
summary(data$SEXSIMPLE)
```

Scenario Calculations:
```{r}
# Calculate the odds ratios for people with different scenarios
# By looking at the mean values for each variable, we can create sample scenarios and compare the likelihood of two people voting

# -2.164897 + (0.077789*AGE) + (0.213431*METROSIMPLECentralCity) + (0.215372*METROSIMPLEOutsideCentralCity) + (0.165660*RACESIMPLEBlack) + (-0.542498*RACESIMPLEAmericanIndian) + (-0.354319*RACESIMPLEAsian) + (-0.168815*RACESIMPLEOther)+ (0.219261*SEXSIMPLEFemale)

b0 = -2.164897 
b_age = 0.077789
b_centCity = 0.213431
b_outCenCity = 0.215372
b_black= 0.165660
b_amer = -0.542498
b_asian = -0.354319
b_other = -0.168815
b_female = 0.219261

# odds for 18 year old, Black, Central City, Female
odds_1 = exp(b0 + b_age*18 + b_centCity*1 + b_black*1
+ b_female*1)
odds_1

# odds for 22 year old, White, Outside Central City, Male
odds_2 = exp(b0 + b_age*22 + b_outCenCity*1)
odds_2

odds_ratio1 = odds_1/odds_2
odds_ratio1
```
An 18 year old, black, central city, female will have an odds of 0.84676 of voting in an election versus a 0.78807 odds of voting for a 22 year old, white, outside central city, male. Therefore, an 18 year old with the characteristics mentioned has 107.44% the odds of a 22 year old.  
```{r}
# odds for 19 year old, Asian, Non-metro area, Female
odds_3 <- exp(b0 + b_age*19 + b_asian*1 + b_female*1)
odds_3

# odds for 20 year old, Black, Central City, Male
odds_4 <- exp(b0 + b_age*20 + b_black*1 + b_centCity*1)
odds_4

odds_ratio2 <- odds_3/odds_4
odds_ratio2
```
An 19 year old, asian, non-metro, female will have an odds of 0.43957 of voting in an election versus a 0.79451 odds of voting for a 20 year old, black, central City, male. Therefore, an 18 year old with the characteristics mentioned has a 55.33% the odds of a 20 year old.

```{r}
# odds for 22 year old, Native American, Non-metro Area, Male
odds_5 <- exp(b0 + b_age*22 + b_amer*1)
odds_5

# odds for 20 year old, White, Non-metro Area, Male
odds_6 <- exp(b0 + b_age*20)
odds_6

odds_ratio3 <- odds_5/odds_6
odds_ratio3
```
It is very interesting because a student who is Native American and is a senior has a 0.3693 odds of voting versus a White 20 year old student with a chance of 0.5438. This shows that race is a major contributor to voter turnout because even though the Native American student is older than the white student, their likelihood of voting is still smaller. 
=======
---
title: "College Voting"
output:
  pdf_document: default
  html_notebook: default
---

```{r}
#Load the necessary libraries
library(ipumsr)
library(tidyverse)
library(readxl)
library(scales)
```


```{r}
#Read in data file and format to grid view
ddi <- read_ipums_ddi("C:/Users/Rinija/Documents/Other/Projects/college-voting/TiS FDC 2020/TiS FDC 2020/data_fixed_width_for_R/cps_00008.xml")
data <- read_ipums_micro(ddi)
View(data)
```

```{r, eval=FALSE}
#Filter data so that only people that are registered to vote included
#Filter data so that only people who "Voted" or "Did not vote" were included
data <- data %>%
  filter(is.na(VOYNOTREG) | VOYNOTREG != 8)  %>%
  filter(VOTED == 1 | VOTED == 2 ) %>%
  filter(!(VOTED == 1 & VOREG == 98))
```

```{r}
#Remove NA entries and re-factor the categorical values  
data <- data %>%
  
  #Metro data into 3 categories
  mutate(
    METROSIMPLE = case_when(
      METRO == 1 ~ 0,  #Not in metro area
      METRO == 2 ~ 1,  #Central city
      METRO == 3 ~ 2   #Outside central city
        )
    )%>% 
  #Get rid of NA values
  filter(!is.na(METROSIMPLE)) %>% 
  #Rename the metro categories
  mutate(METROSIMPLE = fct_recode(as.character(METROSIMPLE), 
                                "Non-metro area" = "0",
                                 "Central city" = "1", 
                                 "Outside central city" = "2"
                         
                            )) %>% 
  #Race data into 4 main categories and others are in the 5th category
  mutate(
    RACESIMPLE = case_when(
      RACE == 100 ~ 1, #White
      RACE == 200 ~ 2, #Black
      RACE == 300 ~ 3, #American Indian/Aleut/Eskimo
      RACE == 651 ~ 4, #Asian only
      RACE > 651 ~ 5,  #Other
      RACE < 830 ~ 5   #Unspecified
         )
     ) %>% 
  #Rename the race categories
  mutate(RACESIMPLE = recode_factor(as.character(RACESIMPLE), 
                                "1" = "White",
                                 "2" = "Black", 
                                 "3" = "American Indian",
                                "4" = "Asian",
                                "5" = "others" )) %>% 
  #Remove NIU variable
  filter(SEX == 1 | #male
         SEX == 2   #female
         ) %>% 
  #Rename sex categories
  mutate(SEXSIMPLE = recode_factor(as.character(SEX), "1" = "Male","2" = "Female")) %>% 
  
  #Re-factor 
  mutate(
    VOTEDSIMPLE = case_when(
      VOTED == 1 ~ 0, #Did not vote
      VOTED == 2 ~ 1  #Voted
        )
  )  %>% 
  
  #Only isolate college students
  filter(AGE > 17, AGE < 23) %>% 
  filter(!is.na(VOTEDSIMPLE)) %>% 
  mutate(VOTEDSIMPLE = recode_factor(VOTEDSIMPLE, "0" = "No Vote","1" = "Vote")
         )
```

Preliminary Analysis of Variable Distributions:

Independent & Dependent Variables ->
  Y-variable - Voted
  X-variables - Age, Metro, Race, SEX
```{r}
#Preliminary Analysis of VOTEHOWSIMPLE
summary(data$VOTEDSIMPLE)
#hist(data$VOTEDSIMPLE)
table(data$VOTEDSIMPLE)
```
VOTEDSIMPLE Data Analysis:

The majority of the population voted and the second highest concentration did not vote. There is roughly half as many people that did not vote when compared to the people that did vote. 
  
```{r}
#Preliminary Analysis of AGE
summary(data$AGE)
hist(data$AGE)
boxplot(data$AGE)
```
Age Data Analysis:

The highest concentration of people is between ages of 55-60
It is worth noting that there were fewer people in this study that were between the ages of 18-20 and ages 70-85. 

```{r}
#Preliminary Analysis of METRO
summary(data$METROSIMPLE)
boxplot(data$METROSIMPLE)
table(data$METROSIMPLE)
```
Metro Data Analysis:

The majority of people in this data set live outside central city and the second highest category is people that live in central cities.


```{r}
#Preliminary Analysis of Race
summary(data$RACE)
hist(data$RACE)
table(data$RACE)
```
Race Data Analysis:

The majority of people in this data set are white. The number of minorities is much smaller compared the the number of white voters. 

```{r}
#Preliminary Analysis of Veterans
summary(data$VETSTAT)
hist(data$VETSTAT)
table(data$VETSTAT)
```
Veterans Data Analysis:
The majority of people in this data set did not serve veterans and only 0.102434 people served.  

```{r}
#Plot voting rates in terms of where people lived 
data %>% select(VOTEDSIMPLE,AGE,RACE,METROSIMPLE,SEXSIMPLE) %>% 
 
  ggplot()+
  geom_bar(aes(x=VOTEDSIMPLE))+
  
  facet_wrap(~METROSIMPLE)+
  scale_y_continuous(label = scientific_format(digits = 2))+
  labs(title = "Voting Rate Based on Location")
```



```{r}
#Plot voting rates in terms of people's race
data %>% 
  select(VOTEDSIMPLE,AGE,RACESIMPLE,METROSIMPLE,SEXSIMPLE) %>% 
  ggplot()+
  #, fill = SEXSIMPLE
  geom_bar(aes(x=VOTEDSIMPLE)) +
  facet_grid(~RACESIMPLE)+

  labs(title = "Voting Rate Based on Race")
```



```{r}
#Plot voting rates in terms of sex
data %>% 
  select(VOTEDSIMPLE,AGE,RACE,METROSIMPLE,SEXSIMPLE) %>% 
 
  ggplot()+
  geom_bar(aes(x=SEXSIMPLE, fill=VOTEDSIMPLE))+
  labs(title = "Voting Rate Based on Sex")
```



```{r}
#Plot voting rate depending on age of college students
data %>% 
  select(VOTEDSIMPLE,AGE,RACE,METROSIMPLE,SEXSIMPLE) %>% 

  ggplot()+
  geom_bar(aes(x=AGE, fill=VOTEDSIMPLE))+
  labs(title = "Voting Rate Based on Age")
```



Logistic Regression:

```{r}
summary(glm(VOTEDSIMPLE ~ AGE + METROSIMPLE + RACESIMPLE + SEXSIMPLE, family = binomial(link = "logit"), data = data))
```
Interpretation of Logistic Regression Model:

Assumptions: Each variable does not effect the other ones. This means that race will not have an effect on metro, sex, or age. 

Model equation:
  log_odds =  -2.164897 + (0.077789*AGE) + (0.213431*METROSIMPLECentralCity) + (0.215372*METROSIMPLEOutsideCentralCity) + (0.165660*RACESIMPLEBlack) + (-0.542498*RACESIMPLEAmericanIndian) + (-0.354319*RACESIMPLEAsian) + (-0.168815*RACESIMPLEOther)+ (0.0932384*SEXSIMPLEFemale)

Hypothesis Test:

  Null Hypothesis -> AGE, RACE, METRO, and SEX did not impact VOTED
  Alternate Hypothesis -> Either AGE, RACE, METRO, or SEX impacts VOTED
  
  Since each variable has a p-value that is less than 0.05, each variable has a significant impact on whether that person votes or not.
  
Values in transformed data:
```{r}
summary(data$AGE)
summary(data$METROSIMPLE)
summary(data$VOTEDSIMPLE)
summary(data$RACESIMPLE)
summary(data$SEXSIMPLE)
```

Scenario Calculations:
```{r}
#Calculate the odds ratios for people with different scenarios
#By looking at the mean values for each variable, we can create sample scenarios and compare the likelihood of two people voting
#Since RACE has the highest impact on the outcome, that will be the changing variable in this example

#-0.5678881 + (0.0260186*AGE) + (0.0909261*METROSIMPLE) + (-0.1168276*RACESIMPLE) + (0.0932384*SEXSIMPLE)

b0 = -0.5678881 
b1 = 0.0260186
b2 = 0.0909261
b3 = -0.1168276
b4 = 0.0932384 

# odds for 18 year old, White, Central City, Female
age <- 18 
metro <- 1
race <- 1
sex <- 1 

log_odds_1 <- b0 + b1*age + b2*metro + b3*race + b4*sex
odds_1 <- exp(log_odds_1)
odds_1

# odds for 18 year old, Black, Central City, Female
age <- 18 
metro <- 1
race <- 2
sex <- 1 

log_odds_2 <- b0 + b1*age + b2*metro + b3*race + b4*sex
odds_2 <- exp(log_odds_2)
odds_2

odds_ratio <- odds_1/odds_2
odds_ratio
```

```{r}
#Finding Odds Ratio for AGE

#-0.5678881 + (0.0260186*AGE) + (0.0909261*METROSIMPLE) + (-0.1168276*RACESIMPLE) + (0.0932384*SEXSIMPLE)

b0 = -0.5678881 
b1 = 0.0260186
b2 = 0.0909261
b3 = -0.1168276
b4 = 0.0932384 

# odds for 19 year old, White, Central City, Female
age <- 19 
metro <- 1
race <- 1
sex <- 1 

log_odds_1 <- b0 + b1*age + b2*metro + b3*race + b4*sex
odds_1 <- exp(log_odds_1)
odds_1

# odds for 18 year old, White, Central City, Female
age <- 18 
metro <- 1
race <- 1
sex <- 1 

log_odds_2 <- b0 + b1*age + b2*metro + b3*race + b4*sex
odds_2 <- exp(log_odds_2)
odds_2

odds_ratio <- odds_1/odds_2
odds_ratio
```

```{r}
#Finding Odds Ratio for METROSIMPLE

#-0.5678881 + (0.0260186*AGE) + (0.0909261*METROSIMPLE) + (-0.1168276*RACESIMPLE) + (0.0932384*SEXSIMPLE)

b0 = -0.5678881 
b1 = 0.0260186
b2 = 0.0909261
b3 = -0.1168276
b4 = 0.0932384 

# odds for 18 year old, White, Central City, Female
age <- 18 
metro <- 1
race <- 1
sex <- 1 

log_odds_1 <- b0 + b1*age + b2*metro + b3*race + b4*sex
odds_1 <- exp(log_odds_1)
odds_1

# odds for 18 year old, White, Outside Central City, Female
age <- 18 
metro <- 2
race <- 1
sex <- 1 

log_odds_2 <- b0 + b1*age + b2*metro + b3*race + b4*sex
odds_2 <- exp(log_odds_2)
odds_2

odds_ratio <- odds_1/odds_2
odds_ratio
```

```{r}
#Finding Odds Ratio for SEX

#-0.5678881 + (0.0260186*AGE) + (0.0909261*METROSIMPLE) + (-0.1168276*RACESIMPLE) + (0.0932384*SEXSIMPLE)

b0 = -0.5678881 
b1 = 0.0260186
b2 = 0.0909261
b3 = -0.1168276
b4 = 0.0932384 

# odds for 18 year old, White, Central City, Female
age <- 18 
metro <- 1
race <- 1
sex <- 1 

log_odds_1 <- b0 + b1*age + b2*metro + b3*race + b4*sex
odds_1 <- exp(log_odds_1)
odds_1

# odds for 18 year old, White, Central City, Male
age <- 18 
metro <- 1
race <- 1
sex <- 0 

log_odds_2 <- b0 + b1*age + b2*metro + b3*race + b4*sex
odds_2 <- exp(log_odds_2)
odds_2

odds_ratio <- odds_1/odds_2
odds_ratio
```
>>>>>>> 8f005c70f6487373f15a0136249375ee5445d5f0
